---
title: "Weather and Corn Yield Regressions"
author: "Laney Gordon"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(R.matlab)
library(rnassqs)
```

## Weather Data Analysis

### Load the PRISM daily maximum temperatures

```{r tmax data}

# daily max temperature
# dimensions: counties x days x years
prism <- readMat("Data/prismiowa (5).mat")

# look at county #1
t_1981_c1 <- prism$tmaxdaily.iowa[1,,1]
t_1981_c1[366]
plot(1:366, t_1981_c1, type = "l")

ggplot() +
  geom_line(mapping = aes(x=1:366, y = t_1981_c1)) +
  theme_bw() +
  xlab("day of year") +
  ylab("daily maximum temperature (°C)") +
  ggtitle("Daily Maximum Temperature, Iowa County #1")


```

```{r tidying up}

# assign dimension names to tmax matrix
dimnames(prism$tmaxdaily.iowa) <- list(prism$COUNTYFP, 1:366, prism$years)

# converted 3d matrix into a data frame
tmaxdf <- as.data.frame.table(prism$tmaxdaily.iowa)

# relabel the columns
colnames(tmaxdf) <- c("countyfp","doy","year","tmax")
tmaxdf <- tibble(tmaxdf)

```

## Temperature trends

### Summer temperature trends: Winneshiek County

```{r temp trends}

tmaxdf$doy <- as.numeric(tmaxdf$doy)
tmaxdf$year <- as.numeric(as.character(tmaxdf$year))

winnesummer <- tmaxdf %>%
  filter(countyfp==191 & doy >= 152 & doy <= 243) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnesummer, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)

lm_summertmax <- lm(meantmax ~ year, winnesummer)
summary(lm_summertmax)

```

### Winter Temperatures - Winneshiek County

```{r winter temps}

winnewinter <- tmaxdf %>%
  filter(countyfp==191 & (doy <= 59 | doy >= 335) & !is.na(tmax)) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnewinter, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)

lm_wintertmax <- lm(meantmax ~ year, winnewinter)
summary(lm_wintertmax)

```

### Multiple regression -- Quadratic time trend

```{r quadratic temp trend}

winnewinter$yearsq <- winnewinter$year^2

lm_wintertmaxquad <- lm(meantmax ~ year + yearsq, winnewinter)
summary(lm_wintertmaxquad)
winnewinter$fitted <- lm_wintertmaxquad$fitted.values

ggplot(winnewinter) +
  geom_point(mapping = aes(x = year, y = meantmax)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  labs(x = "year", y = "tmax") 

```

### Download NASS corn yield data

```{r download, results= 'hide' } 

# set our API key with NASS
nassqs_auth(key = "BC58E27D-5B67-3236-A9E4-8E1CE4BBE26F")

# parameters to query on 
params <- list(commodity_desc = "CORN", util_practice_desc = "GRAIN", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
cornyieldsall <- nassqs_yields(params) 

cornyieldsall$county_ansi <- as.numeric(cornyieldsall$county_ansi)
cornyieldsall$yield <- as.numeric(cornyieldsall$Value)

# clean and filter this dataset
cornyields <- select(cornyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
cornyields <- tibble(cornyields) 

```

## Assignment

### Question 1a: Extract Winneshiek County corn yields, fit a linear time trend, make a plot. Is there a significant time trend?
```{r}

winnecornyields <- cornyields %>%
  filter(county_ansi=="191") 

ggplot(winnecornyields, mapping = aes(x= year, y= yield)) +  
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "yield") +
  geom_smooth(method = lm) 

lm_winnecornyields <- lm(yield ~ year, winnecornyields)
summary(lm_winnecornyields) 
```
  
There is a positive trend displayed from the Winneshiek County corn yields plot that shows an increase in yield as time increases.The p-value of the data (1.767e^-13) is smaller than alpha (0.05), also telling us there is significant evidence that a time trend is present.   

### Question 1b: Fit a quadratic time trend (i.e., year + year^2) and make a plot. Is there evidence for slowing yield growth? 
```{r}
winnecornyields$yearsq <- winnecornyields$year^2

lm_cornyieldsquad <- lm(yield ~ year + yearsq, winnecornyields)
summary(lm_cornyieldsquad) 
winnecornyields$fitted <- lm_cornyieldsquad$fitted.values

ggplot(winnecornyields) +
  geom_point(mapping = aes(x = year, y = yield))+
  geom_line(mapping = aes(x = year, y = fitted)) + 
  theme_bw() +
  labs(x = "year", y = "yield") 

```
  
There is significant evidence that there is a relationship between yield growth and year due to the p-value being smaller than alpha, however the graph displays a positive slope, thus the evidence is NOT slowing yield growth.  

### Question 2 -- Time Series: Let's analyze the relationship between temperature and yields for the Winneshiek County time series. Use data on yield and summer avg Tmax. Is adding year or Tmax^2 to your model helpful? Make a plot and interpret the results.

```{r} 
#combining summer temp with yield
temp_yield <- inner_join(winnecornyields, winnesummer) 

#quadratic temp
temp_yield$meantmaxsq <- temp_yield$meantmax^2 

lm_temp_yield <- lm(yield ~ meantmax + meantmaxsq, temp_yield)
summary(lm_temp_yield) 

temp_yield$quadfitted <- lm_temp_yield$fitted.values

#plot
ggplot(temp_yield) +
  geom_point(mapping = aes(x = meantmax, y = yield))+
  geom_line(mapping = aes(x = meantmax, y = quadfitted)) + 
  theme_bw() +
  labs(x = "Temp (C)", y = "Yield") 
```

Adding Tmax^2 to the model is beneficial to the data because it helps us identify the correlation between temperature and yield. The graph displays normal distribution from temperatures 24 degrees C to about 31 degrees C and shows the highest yield occurs from temperatures 26 to 28 degrees C. The p-value of 0.0078 is less than alpha (0.05), thus we have evidence that the finding is significant evidence.  


### Question 3 -- Cross-Section: Analyze the relationship between temperature and yield across all counties in 2018. Is there a relationship? Interpret the results.


```{r}
# Want yield for all counties, 2018:
county_yield <- cornyields %>% 
  filter(year == 2018) %>% 
  group_by(county_name) %>% 
  unique() %>% 
  filter(!is.na(county_ansi))


#Want maximum summer temperatures (presumably) for 2018, per county. 
maxtemp18 <- tmaxdf %>%
  group_by(countyfp) %>%
  filter(year == 2018) %>% 
  filter(doy >= 152 & doy <= 243) %>%
  rename(county_ansi = "countyfp") %>%
  summarize(meantmax = mean(tmax))  
  

#change from factor to numeric

maxtemp18$county_ansi <- as.numeric(as.character(maxtemp18$county_ansi))
county_yield$county_ansi <- as.numeric(as.character(county_yield$county_ansi))

#Joining the two together
combo <- inner_join(maxtemp18, county_yield, by='county_ansi')

combo$meantmaxsq <- combo$meantmax^2  

#linear model
lm_Q3 <- lm(yield ~ meantmax + meantmaxsq, data = combo) 
summary(lm_Q3) 

#quadratic
combo$fitted <- lm_Q3$fitted.values 

#plot
ggplot(combo) + 
   geom_point(mapping = aes( x = meantmax, y= yield)) + 
  geom_line(mapping = aes( x= meantmax, y = fitted)) +
  theme_bw() +
  labs(x = "Avg Max Temperature (C)", y = "Yield") 

```
  
Based on the distribution of values displayed in our plot, there is not a clear distinction of a relationship between meantmax and yield across all Iowa counties in 2018. There is enough significant evidence to make this claim because the p-value of 0.0017 is less than alpha, which reassures us that we reject the null hypothesis that there is a relationship between the two.  


### Question 4 -- Panel: One way to leverage multiple time series is to group all data into what is called a "panel" regression. Convert the county ID code ("countyfp" or "county_ansi") into factor using as.factor, then include this variable in a regression using all counties' yield and summer temperature data. How does the significance of your temperature coefficients (Tmax, Tmax^2) change? Make a plot comparing actual and fitted yields and interpret the results of your model.

```{r}
#making the county_ansi into a factor:
cornyields$county_ansi <- as.factor(as.character(cornyields$county_ansi)) 


#confirming it is a factor
str(cornyields$county_ansi) 

#summer temperatures
summertemp <- tmaxdf %>%
  group_by(countyfp, year) %>%
  filter(doy >= 152 & doy <= 243) %>%
  summarize(meantmax = mean(tmax)) %>% 
  rename(county_ansi = "countyfp") 

summerq4 <- inner_join( summertemp, cornyields) %>%
  unique() 
 
# Squares the tmax value
summerq4$meantmaxsq <- summerq4$meantmax^2

lm_Q4 <- lm(yield ~ county_ansi + meantmax + meantmaxsq + year, summerq4)
summary(lm_Q4) 

summerq4$fitted <- lm_Q4$fitted.values 

#plot
ggplot(summerq4, aes( x = fitted, y= yield)) +
   geom_point() + 
  theme_bw() +
  labs(x = "Fitted Yield", y = "Actual Yield") +  
  geom_abline(color= "green", size = 1)   
```

As the actual yield values increase, the fitted yields also increase, and vice versa, creating a diagonal pattern. This diagonal pattern reassures us that the predicted values were fairly close to the actual yield values and there were no obvious outliers.The p-value was extremely low, meaning we have significant evidence of this relationship. 


### Question 5 -- Soybeans: Download NASS data on soybean yields and explore either a time series relationship for a given county, the cross-sectional relationship for a given year, or a panel across all counties and years.
```{r yield download, results= 'hide' }

# set our API key with NASS
nassqs_auth(key = "BC58E27D-5B67-3236-A9E4-8E1CE4BBE26F")

# parameters to query on 
params2 <- list(commodity_desc = "SOYBEANS", statisticcat_desc = "YIELD", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
soybeansall <- nassqs_yields(params2)  

soybeansall$county_ansi <- as.numeric(soybeansall$county_ansi) 
soybeansall$yield <- as.numeric(soybeansall$Value) 

# clean and filter this dataset
soybeanyields <- select(soybeansall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
soybeanyields <- tibble(soybeanyields) 
```

```{r soy yield}

#specific Winneshiek County
winnesoy_yield <- soybeanyields %>%
  filter(county_ansi=="191") 


#time trend
ggplot(winnesoy_yield, mapping = aes(x= year, y= yield)) +  
  geom_point() +
  theme_bw() +
  labs(x = "Year", y = "Yield") +
  ggtitle('Soybean Yield Across Winneshiek County') + 
  geom_smooth(method = lm) 

lm_winnesoy_yield <- lm(yield ~ year, winnesoy_yield)
summary(lm_winnesoy_yield) 

```


### Bonus: Find a package to make a county map of Iowa displaying some sort of information about yields or weather. Interpret your map.

```{r iowa map}
#download packages
library(ggplot2)
library(maps) 
library(mapdata)

#import us map
usa <- map_data('usa') 
ggplot(data=usa, aes(x=long, y=lat, group=group)) + 
  geom_polygon(fill='lightblue') + 
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) + 
  ggtitle('U.S. Map') + 
  coord_fixed(1.3) 

#state outlines
state <- map_data("state")
ggplot(data=state, aes(x=long, y=lat, fill=region, group=group)) + 
  geom_polygon(color = "white") + 
  guides(fill= "none") + 
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) + 
  ggtitle('U.S. Map with States') + 
  coord_fixed(1.3)

#Iowa
iowa <- subset(state, region=="iowa")
counties <- map_data("county")
iowa_county <- subset(counties, region=="iowa")

ca_map <- ggplot(data=iowa, mapping=aes(x=long, y=lat, group=group)) + 
  coord_fixed(1.3) + 
  geom_polygon(color="black", fill="lightgreen") + 
  geom_polygon(data=iowa_county, fill=NA, color="white") + 
  geom_polygon(color="black", fill=NA) + 
  ggtitle('Iowa Map with Counties') + 
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank())
ca_map

```
  
This code was found on [github](https://jtr13.github.io/cc19/different-ways-of-plotting-u-s-map-in-r.html) . I was unable to incorporate a database for weather or yield, however I believe learning to graph state maps is extremely useful. 

### Bonus #2: Challenge question - map trends in corn yields by county across Iowa. Interpret your map.
